---
title: "Assignment 3"
author: "Tirza IJpma, Mick IJzer and Maud van den Berg"
date: "3/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
a) Study the data and give a few (> 1) summaries (graphics or tables).
```{r, echo = F}
data = read.table("psi.txt", header=T)
tot=xtabs(~passed+psi+round(gpa),data=data)
attach(data)
data
par(mfrow = c(1,3))
hist(passed)
hist(psi)
hist(gpa)
```
b) Fit a logistic regression model with both explanatory variables. Does psi work?
```{r}
data$psi = as.factor(psi)
attach(data)
dataglm = glm(passed~psi+gpa, data = data, family=binomial)
summary(dataglm)
```

c) Estimate the probability that a student with a gpa equal to 3 who receives psi passes the assignment. Estimate the same probability for a student who does not receive psi.
```{r}
 predict(dataglm, newdata=data.frame(gpa=c(3,3), psi=factor(c(0,1))), type= 'response')

```
         1          2 
0.08230274 0.48158645 --> 1: gpa of 3 and no psi, 2: gpa of 3 and psi.

d) Estimate the relative change in odds of passing the assignment rendered by instructing students with psi rather than the standard method (for an arbitrary student). What is the interpretation of this number? Is it dependent on gpa?

```{r}

```

